% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      

% \textit{“It takes a village (to raise a child).”}\\
% \rightline{African proverb}

% \vspace{5mm}

\par First and foremost, I would like to acknowledge and express my sincere gratitude to Maike Sander, Birgit Sawitzki and Sascha Sauer for providing me with this wonderful opportunity to undertake my doctoral work under their supervision. Your warm and generous guidance has taught me so much about the ebbs and flows of research, collaborations, and what being a researcher is all about. I deeply appreciate the unwavering support you provided me with especially when times were challenging in my personal life, and for that, I shall forever be grateful.\\

\par To Altuna, thank you for making me part of your group, for giving me the freedom to develop this work and for providing me with valuable feedback. To Thomas, thank you for stepping in whenever I needed a critical eye and for helping me find motivation when research got mundane. To my Thesis Advisory Committee - Laleh and Sigmar, thank you for regularly checking in me about the progress of my work and for ensuring that the thesis did not veer too far off course.\\

\par To Matt, Han, Matthias, Ibrahim, Kerstin and Maria, all of whom have guided me on a daily basis throughout the course of this thesis. Thank you as you bore with me through the daily lows and infrequent highs of scientific research. In particular, I owe a debt of gratitude to Matt and Han for our weekly discussions about the projects, sometimes\\

\par To MDC, Charité, BIH and Freie Universität Berlin, thank you for providing an outstanding environment and to perform my research.\\

\par To Kerim, Sarah, and Iza, thank you for everything! As clichéd as it may sound, words truly fall short in expressing the immense love and warmth I've felt from the three of you over the past six years. You have been there from the very beginning of my doctoral journey, and together we have shared countless moments of laughter, joy, sadness, heartbreak, burnout, and depression. Through thick and thin, you've stood by me, no questions asked! Your support was unwavering, especially during times when I was difficult to deal with. Our evening coffee breaks have been a sanctuary where I could be myself and momentarily escape the rigors of academia. The weekend hangouts, birthday celebrations, and both planned and impromptu gatherings will forever hold a special place in my heart. As Dom Toretto would say, \textit{``We are Family''}. To many more adventures!   \\

\par To Tijana, \\

\par To all the lovely members of the Ohler lab: Pia, Ece, Henrik, Pascal, Izabel, Vika and others, thank you for adopting me into your group and your social circle. To many more brunches, picnics and celebrations!\\

\par To all the members of Maike's old lab at UCSD and her new lab at MDC, members of Birgit's lab at BIH, members of Scientific Genomics Platforms and members of Altuna's group, thank you for all the critical inputs and supportive feedback you've provided during the several lab meetings.\\

\par To Karthik Nair, my \textit{cheta}, thank you for being patient with me as I rambled my way though numerous heartbreaks and struggles and pain over the last decade. Your inquisitiveness and unfiltered passion for so many things is something I have drawn inspiration from numerous times in the last years.\\

\par To my friends across the Atlantic, thank you for putting up with my late night (your mid-day) complaints about my thesis.\\


\par To all the wonderful people I've crossed paths with over the years and befriended along the way — Sri, Prakhar, Solenne, Lina, Olga, Kasia, Fede, Sam, Selin, Laura, Sara and Tomi, thank you for all the great memories. I will cherish every single one of them.\\

\par To 

\par And, finally, to Berlin, you have my heart.




%
% I cannot emphasise enough how much I have learnt in the last four years, and how much I have grown, as a person and as a scientist.
% I am thankful to EMBL-EBI and the University of Cambridge for providing an outstanding environment to perform my research - I have been truly spoilt.
% I am thankful to my thesis advisory committee, Eileen Furlong and Duncan Odom, and to my collaborators, Ludovic Vallier, Mariya Chhatriwala, Daniel Gaffney, Natsuhiko Kumasaka and Florian Merkle. \\

% The help and guidance of several people was instrumental in making my PhD productive and exciting.
% A special thank you goes to Marc Jan Bonder and Daniel Seaton, for being the most amazing colleagues and for their continued presence and support, at work and outside, especially in the second part of my PhD. 
% Thank you for the morning coffees, the board games, and just for always being there for me. \\

% I would also like to thank Paolo Casale, who in the first few months of my PhD introduced and explained to me key concepts within the field of Statistical Genetics; Davis McCarthy for introducing me to and patiently helping me with all things R and single cell RNA-seq related; and Nils Eling for helping me navigate life as an EBI and Cambridge PhD student in those early days.
% I would also like to thank Danilo Horta, for his availability to answer all my questions to do with software and maths.
% I would like to thank Julie Jerber, Ximena Ibarra-Soria, Arianne Richard, Rebecca Berrens, Na Cai, Rachel Moore and Sara Pulit for teaching me all the biology and genetics I know, and for being fantastic role models for me, whether they are aware or not!
% I am especially thankful to Julie with whom I have worked on one of the projects during my PhD, it was a truly wonderful collaboration and I could not have wished for a better partner.
% I would also like to thank all other past and present members of the Marioni group, particularly Aaron, Chris, Mike, Rebecca, Shila and Tom. \\

% I would like to thank the following people for reading and commenting this thesis: Shila Ghazanfar, Nils Eling, Julie Jerber, Daniel Seaton, Michael Morgan, Ximena Ibarra-Soria, Paula Frampton, Kendall MacMillan, Oliver Stegle and John Marioni.

% I would like to thank Nils Eling and Krishna Kumar for this amazing template. 
% I would also like to thank my viva examiners Professor Cecilia Lindgren and Dr. Chris Wallace for reading my thesis in detail and for fruitful discussion and comments. \\

% Finally, I would like to thank Hannah Currant and Elissavet (Elsa) Kentepozidou, for being the best companions in this PhD journey.
% From the very first days in Heidelberg, to wedding dress shopping, to discussing work, science, equality, pandemics and everything in between, I could not have asked for two better people to do this thing with.
% I would also like to thank the rest of the amazing community that are the EBI predocs, including the rest of my batch, besides Hannah and Elsa - Ricard, Umberto, Harald and Melike; as well as previous batches especially Nils, Jack, Dani, Hannah, Nadia and Claudia; and following batches, including Rachel, Martina, Conor, Jose. 
% Special mention also to Pauline and Joanna, from EMBL Grenoble. \\

% On a more personal note, I would like to thank my friends and family, who despite the distance are always there for me, and are my biggest cheerleaders.
% I would like to thank Livia, Teresa, Camilla, Sofia and Martina for having been by my side for the last 22+ years.
% Giulia for always believing in me and for being with me on the most important day.
% Nagiua and Michele, and the rest of the (extended) family, for everything.
% I would like to thank my parents, Daniela and Renato, for always encouraging my curiosity and for making me who I am today.
% “Grazie mamma, grazie pa!”, cit.
% Last but not least, I would like to thank Kendall, for always believing in me and for being the best partner anyone could ask for.

\end{acknowledgements}