\chapter{List of Publications}
\label{chp:publications}

\clearpage
%\newpage\null\thispagestyle{empty}\newpage

% Change hyperlink color to blue
\hypersetup{urlcolor=blue, linkcolor=blue, citecolor=blue}

\subsubsection{\Large Published papers}

Friedel, Caroline C. and Whisnant, Adam W. and Djakovic, Lara and Rutkowski, Andrzej J. and Friedl, Marie-Sophie and Kluge, Michael and Williamson, James C. and \underline{Sai, Somesh} and Vidal, Ramon Oliveira and Sauer, Sascha and Hennig, Thomas and Grothey, Arnhild and Milić, Andrea and Prusty, Bhupesh K. and Lehner, Paul J. and Matheson, Nicholas J. and Erhard, Florian and Dölken, Lars. \textbf{Dissecting Herpes Simplex Virus 1-Induced Host Shutoff at the RNA Level.} \textit{Journal of Virology} \textbf{95} and (2021) DOI: \href{https://doi.org/10.1128/jvi.01399-20}{https://doi.org/10.1128/jvi.01399-20}\\

\subsubsection{\Large Accepted Manuscripts}

Bakina, Olga and Conrad, Thomas and Mitic, Nina and Vidal, Ramon Oliveira and Obrusnik, Tessa and \underline{Sai, Somesh} and Nolte, Christiane and Semtner, Marcus and Kettenmann, Helmut. \textbf{In situ Patch-seq analysis of microglia reveals a lack of stress genes as found in FACS-isolated microglia.} \textit{bioRxiv} DOI: \href{10.1101/2023.03.22.533782}{10.1101/2023.03.22.533782}\\

\subsubsection{\Large Submitted Manuscripts}

Wortham, Matthew\textsuperscript{*} and Ramms, Bastian\textsuperscript{*} and Zeng, Chun\textsuperscript{*} and Benthuysen, Jacqueline R.\textsuperscript{*} and \underline{Sai, Somesh} and Pollow, Dennis P. and Liu, Fenfen and Schlichting, Michael and Harrington, Austin R. and Liu, Bradley  and Prakash, Thazha P. and Pirie, Elaine C and Zhu, Han and Baghdasarian, Siyouneh and Auwerx, Johan and Shirihai, Orian S. and Sander, Maike. \textbf{Metabolic control of adaptive $\beta$-cell proliferation by the protein deacetylase SIRT2.} \textit{bioRxiv} DOI: \href{10.1101/2024.02.24.581864}{10.1101/2024.02.24.581864}\\\textit{\textsuperscript{*} equal contributions}\\

Nguyen-Ngoc, Kim-Vy\textsuperscript{*} and Jun, Yesl\textsuperscript{*} and \underline{Sai, Somesh} and R. Bender, Hugh F. and Kravets, Vira and Zhu, Han and Hatch, Christopher J. and Schlichting, Michael  and Gaetani, Roberto and Mallick, Medhavi and Hachey, Stephanie J. and Christman, Karen L.  and George, Steven C.  and Hughes, Christopher C.W. and Sander, Maike. \textbf{Engineered vasculature induces functional maturation of pluripotent stem cellderived islet organoids.} \textit{bioRxiv} DOI: \href{10.1101/2022.10.28.513298}{10.1101/2022.10.28.513298}\\\textit{\textsuperscript{*} equal contributions}\\


\clearpage

Altieri, Barbara\textsuperscript{*} and Secener, A. Kerim\textsuperscript{*} and Sai, Somesh and Fischer, Cornelius and Sbiera, Silviu and Arampatzi, Panagiota and Herterich, Sabine and Landwehr, Laura-Sophie and Vitcetz, Sarah N. and Braeuning , Caroline and Fassnacht, Martin and Ronchi, Cristina L. and Sauer, Sascha. \textbf{Cell Atlas at Single-Nuclei Resolution of the Adult Human Adrenal Gland and Adrenocortical Adenomas.} \textit{bioRxiv} DOI: \href{10.1101/2022.08.27.505530}{10.1101/2022.08.27.505530}\\\textit{\textsuperscript{*} equal contributions}\\


\subsubsection{\Large Manuscripts in preparation}

Zhu, Han\textsuperscript{*} and \underline{Sai, Somesh\textsuperscript{*}} and Omar, Ibrahim\textsuperscript{*} and Barone, Matthias and Mühle, Kerstin and Schneider, Maria and Liu, Fenfen and Sriram, Sharanya and Johnson, Juliette C. and Thoma, Tizia and Sauer, Sascha and Sawitzki, Birgit and Sander, Maike. \textbf{Western diet feeding aggravates age-dependent accumulation of inflammatory macrophages and CD8+ cytotoxic T cells in pancreatic tissue.}\\\textit{\textsuperscript{*} equal contributions}\\

Iwert, Christina\textsuperscript{*} and Stein, Julia\textsuperscript{*} and Appelt, Christine and Vogt, Katrin and Johnson, Juliette C. and Dey, Hannah P. and Oktay, Hulya Zeynep and Steiner, Emma and Mühle, Kerstin and Stanko, Katarina and Uebe, Doreen and  Sell, Thomas and Fritsche-Guenther, Raphaela and Rainer, Roman Josef and Tummler, Katja and Jürchott, Karsten and Gutbier, Birgitt and Nouailles, Geraldine  and Lisec, Jan and Janek, Katharina and Gille, Christoph and Textoris-Taube, Kathrin and Bolaji, Olufemi and Petrov, Lev and \underline{Sai, Somesh} and Kühl, Anja A. and Klipp, Edda and Kirwan, Jennifer A. and Witzenrath, Martin  and Blüthgen, Nils and Meisel, Christian and Sawitzki, Birgit. \textbf{The mitochondrial protein TCAIM acts as a negative regulator of VDAC2 conductance, and T Cell-Mediated Pathology.}\\\textit{\textsuperscript{*} equal contributions}\\





\hypersetup{
    colorlinks=true,
    linkcolor=black,
    filecolor=blue,      
    urlcolor=blue,
}

